%!TEX root = ../swanton-texts.tex
%%
%% 301. Raven’s nose (p. 390)
%%

\resetexcnt
\chapter{Raven’s nose}\label{ch:301-raven-nose}

\ex\label{ex:301-1-}%
\exmn{301.1}%
\begingl
	\glpreamble	Dēł yᴀx wudᴀtsē′n cēyē′ł. //
	\glpreamble	De l yáx̱ wudatseen shéi, Yéil. //
	\gla	De l yáx̱ \rlap{wudatseen} @ {} @ {} @ {} @ {} shéi, Yéil. //
	\glb	de l yáx̱ u- wu- d- \rt[¹]{tsin} -μμL shí Yéil //
	\glc	now \xx{neg} \xx{sim} \xx{irr}- \xx{pfv}- \xx{mid}- \rt[¹]{alive} -\xx{var} \xx{mir} Raven //
	\gld	now not like \rlap{\xx{irr}.\xx{zcnj}.\xx{pfv}.lively} {} {} {} {} surprise Raven //
	\glft	‘Now nothing is as lively (as) Raven.’
		//
\endgl
\xe

\ex\label{ex:301-2-}%
\exmn{301.1}%
\begingl
	\glpreamble	Haỵidē′ wugudī′n. //
	\glpreamble	Haa ÿeedé wugoodéen. //
	\gla	{} Haa \rlap{ÿeedé} @ {} {} \rlap{wugoodéen.} @ {} @ {} @ {} //
	\glb	{} haa ÿee -dé {} wu- \rt[¹]{gut} -μμL -ín //
	\glc	{}[\pr{PP} \xx{1pl·pss} below -\xx{all} {}] \xx{pfv}- \rt[¹]{go·\xx{sg}} -\xx{var} -\xx{past} //
	\gld	{} our below -to {} \rlap{\xx{ncnj}.\xx{pfv}.go·\xx{sg}.\xx{past}} {} {} {} {} //
	\glft	‘He had gone below us.’
		//
\endgl
\xe

\ex\label{ex:301-3-}%
\exmn{301.1}%
\begingl
	\glpreamble	Ag̣ā′gucî dułuwu′ ke wududzîyᴀ′q. //
	\glpreamble	Aag̱áa gushí du loowú kei wududziyíḵ. //
	\gla	{} \rlap{Aag̱áa} @ {} {} gushí {} du \rlap{loowú} @ {} {}
		kei @ \rlap{wududziyíḵ.} @ {} @ {} @ {} @ {} @ {} @ {} //
	\glb	{} á -g̱áa {} gushí {} du lú -í {}
		kei= wu- du- d- s- i- \rt[²]{ÿiḵ} -μH //
	\glc	{}[\pr{PP} \xx{3n} -\xx{ades} {}] \xx{dub} {}[\pr{DP} \xx{3h·pss} nose -\xx{pss} {}]
		up= \xx{pfv}- \xx{4h·s}- \xx{mid}- \xx{xtn}- \xx{stv}- \rt[²]{pull} -\xx{var} //
	\gld	{} then -after {} apparently {} his nose {} {}
		up \rlap{\xx{zcnj}.\xx{pfv}.ppl.fish} {} {} {} {} {} {} //
	\glft	‘Then apparently people fished up his nose.’
		//
\endgl
\xe

\ex\label{ex:301-4-}%
\exmn{301.2}%
\begingl
	\glpreamble	Ag̣a′ āntū′x ỵāwagu′t dułū′wug̣a. //
	\glpreamble	Aag̱áa aan tóox̱ ÿaawagút du loowúg̱aa. //
	\gla	{} \rlap{Aag̱áa} @ {} {} {} aan \rlap{tóox̱} @ {} {}
		\rlap{ÿaawagút} @ {} @ {} @ {} @ {}
		{} du \rlap{loowúg̱aa.} @ {} @ {} {} //
	\glb	{} á -g̱áa {} {} aan tú -x̱ {}
		ÿ- wu- i- \rt[¹]{gut} -μH
		{} du lú -í -g̱áa {} //
	\glc	{}[\pr{PP} \xx{3n} -\xx{ades} {}] {}[\pr{PP} town inside -\xx{pert} {}]
		\xx{qual}- \xx{pfv}- \xx{stv}- \rt[¹]{go·\xx{sg}} -\xx{var}
		{}[\pr{PP} \xx{3h·pss} nose -\xx{pss} -\xx{ades} {}] //
	\gld	{} then -after {} {} town inside -at {}
		\rlap{off.\xx{zcnj}.\xx{pfv}.go·\xx{sg}} {} {} {} {} 
		{} his nose {} \·for {} //
	\glft	‘Then he went along inside of town for his nose.’
		//
\endgl
\xe

\FIXME{obliquely/circuitously \fm{NP-x̱} \fm{ÿa-u-} \~\ \fm{ÿaa=} (\fm{∅}; \fm{-ch} repetitive)}

\ex\label{ex:301-5-}%
\exmn{301.2}%
\begingl
	\glpreamble	ᴀcdjī′t dutī′ ān gānt wudîqî′n. //
	\glpreamble	Ash jeet dutée, aan gáant wudiḵín. //
	\gla	{} {} Ash \rlap{jeet} @ {} {}
			\rlap{dutée,} @ {} @ {} @ {} @ {} {} +
		{} \rlap{aan} @ {} {} {} \rlap{gáant} @ {} {}
		\rlap{wudiḵín.} @ {} @ {} @ {} @ {} //
	\glb	{} {} ash jee -t {}
			{} du- \rt[¹]{ti} -μμH {} {}
		{} á -n {} {} gáan -t {}
		wu- d- i- \rt[¹]{ḵin} -μH //
	\glc	{}[\pr{CP} {}[\pr{PP} \xx{3prx·pss} poss’n -\xx{pnct} {}]
			\xx{zcnj}\· \xx{4h·s}- \rt[²]{handle} -\xx{var} \·\xx{sub} {}]
		{}[\pr{PP} \xx{3n} -\xx{instr} {}] {}[\pr{PP} outside -\xx{pnct} {}]
		\xx{pfv}- \xx{mid}- \xx{stv}- \rt[¹]{fly·\xx{sg}} -\xx{var} //
	\gld	{} {} his poss’n -to {}
			\rlap{\xx{csec}.ppl.handle} {} {} {} \·when {}
		{} it -with {} {} outside -to {}
		\rlap{\xx{zcnj}.\xx{pfv}.fly·\xx{sg}} {} {} {} {} //
	\glft	‘People having given it to him, he flew outside with it.’
		//
\endgl
\xe

\ex\label{ex:301-6-}%
\exmn{301.3}%
\begingl
	\glpreamble	Ayê′x ᴀnsînī′ dog̣odjiỵᴀqaỵî′. //
	\glpreamble	A yáx̱ ansinée du g̱oojí ÿaḵáaÿi. //
	\gla	{} A yáx̱ {} \rlap{ansinée} @ {} @ {} @ {} @ {} @ {}
		{} du \rlap{g̱oojí} @ {} \rlap{ÿaḵáaÿi.} @ {} @ {} {} //
	\glb	{} a yáx̱ {} a- n- s- i- \rt[¹]{niʰ} -μμH
		{} du g̱ooch -í ÿá- ḵáaʷ -í {} //
	\glc	{}[\pr{PP} \xx{3n} \xx{sim} {}] \xx{arg}- \xx{ncnj}- \xx{csv}- \xx{stv}- \rt[¹]{happen} -\xx{var}
		{}[\pr{DP} \xx{3h·pss} wolf -\xx{pss} face- man -\xx{pss} {}] //
	\gld	{} it like {} \rlap{3>3.\xx{rlzn}.make.happen} {} {} {} {} {}
		{} his \rlap{wolf’s} {} \rlap{partner} {} {} {} //
	\glft	‘He has at last made him like it, his wolf partner.’
		//
\endgl
\xe

\FIXME{\fm{ÿaḵáaÿi} not expected \fm{ÿaḵáawu}}

\ex\label{ex:301-7-}%
\exmn{301.3}%
\begingl
	\glpreamble	Ān gānt wudîqî′n. //
	\glpreamble	Aan gáant wudiḵín. //
	\gla	{} \rlap{Aan} @ {} {} {} \rlap{gáant} @ {} {}
		\rlap{wudiḵín.} @ {} @ {} @ {} @ {} //
	\glb	{} á -n {} {} gáan -t {}
		wu- d- i- \rt[¹]{ḵin} -μH //
	\glc	{}[\pr{PP} \xx{3n} -\xx{instr} {}] {}[\pr{PP} outside -\xx{pnct} {}]
		\xx{pfv}- \xx{mid}- \xx{stv}- \rt[¹]{fly·\xx{sg}} -\xx{var} //
	\gld	{} it -with {} {} outside -to {}
		\rlap{\xx{zcnj}.\xx{pfv}.fly·\xx{sg}} {} {} {} {} //
	\glft	‘He flew outside with it.’
		//
\endgl
\xe

\ex\label{ex:301-8-}%
\exmn{301.4}%
\begingl
	\glpreamble	Wâ′sa yūłcîtî′kᵘdaỵa tc!ᴀ nao g̣adᴀnᴀ′. //
	\glpreamble	Wáa sáyú l \{shitíkw daaÿá\}, chʼa naaw g̱adaná. //
	\gla	 //
	\glb	 //
	\glc	 //
	\gld	 //
	\glft	‘’
		//
\endgl
\xe

\ex\label{ex:301-9-}%
\exmn{301.4}%
\begingl
	\glpreamble	Atū′nᴀx nī′tct nᴀ′g̣egut. //
	\glpreamble	A tóonáx̱ néecht nag̱eegoot. //
	\gla	 //
	\glb	 //
	\glc	 //
	\gld	 //
	\glft	‘’
		//
\endgl
\xe