%!TEX root = ../swanton-texts.tex
%%
%% 105. Kwáashkʼi Ḵwáan History (347–368)
%%

\resetexcnt
\chapter{Kwáashkʼi Ḵwáan Daat: Kwáashkʼi Ḵwáan History}\label{ch:105-kwaashkikhwaan-history}

\section{Swanton’s abstract}\label{sec:105-swanton-abstract}

After the death of an Athapascan chief on Copper river the people of his house began fighting over the possession of a dish, and those who lost it determined to emigrate.
They set out, and, after losing some of their people in the mist, crossed the mountains near Mount St.\ Elias and came to the sea.
While they were living at the mouth of a large river a little girl reared a sea gull which attained enormous proportions.
When it got large her brothers wanted to kill it, but it disappeared together with the girl.
By and by thee chief sent six brothers alongshore to hunt for other people.
They encountered head winds at one place and lost one of their number, but finally reached Yakutat.
When they got back they heard that the name of the girl who had reared the sea gull had been given to another.
This girl dug up some thing long and red in the forest out of which a dish was carved like the one that had been taken away from them.
Presently they called in some Burnt-house people who were coming south from the mouth of the Copper river and entertained them.

After this the six brothers were sent back to Copper river for a certain copper plate that had been left there, and they went and came in twenty days each way, half the time it had taken the whole band to come out.
THen the people all set out for Yakutat, where they were at first received in a hostile manner by the Kosk!ē′dî and ʟ!uq!oe′dî living there, until they purchased a creek from them with the copper plate and settled upon it.
This was Kâck!, from which they obtained their name.
By and by the six brothers went hunting, and one of them became so lazy that the others had left him for some time.
Then a mountain being came to him and helped him to become a great hunter.
Finally he sent him home in a canoe which was really a grizzly bear.
This bear turned around to be fed when it was hungry, and that is what made the turns in the river.
After a time the brothers went up to a glacier at the head of Kâck! to hunt, but their canoe was carried away by a swell raised by falling ice.
After waiting in vain for succor, they started to cross the glacier, but one of them became dizzy and was carried away by the Wolf people.
The others got across and were in a starving condition, when the youngest discovered a mountain sheep with very large horns, that was really “the mother of the bears”.
After that they reached home.

By and by the six brothers started south with their brother-in-law.
They tried to cross from Kᴀstaxē′xda-ān to Auk, but were delayed for months by storms.
Finally Heavy-wings, their brother-in-law, discovered that this was caused by North Wind, who was in love with his daughter, and he gave her to him.
After that Heavy-wings saw and caught the ʟ!ê′nᴀxx̣ī′dᴀq and became wealthy, but because he did not use exactly the right words at that time he was killeed by a copper falling upon him and cutting him open.
His nephew also saw the ʟ!ê′nᴀxx̣ī′dᴀq,, used the right words toward it, and became very wealthy.

\section{Swanton’s translation}\label{sec:105-swanton-translation}

\section{Paragraph 1}\label{sec:105-para-1}

%\clearpage
